### 2016-11-14 Paper_v1.tex
Nov14_3

%\usepackage{algorithm2e}
%\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described all of the training data, especially the central carbon metabolism.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
%The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation. Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

Mathematical modeling has long contributed to our understanding of metabolism.
Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction~\cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
We also defined the single set with lowest cost function as our best-fit set.

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).

@@

The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation; and 1 pyrophosphatase.
The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
In addition, a handful of alanine synthesis pairwise sensitivities are high.
However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO$_4$ transport.
This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
We first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig.~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
In these cases the majority of the flux continued through the Entner–Doudoroff pathway, whereas in the unconstrained case the majority of flux traveled through glycolysis.
In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
In the TX/TL constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell-free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
The decrease in acetate production is promising as a mechanism to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

>>>>

%\usepackage{algorithm2e}
%\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}
\hyphenation{dec-ades} %Latex does not know how to hyphenate this by default, so we add it to the list

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described all of the training data, especially the central carbon metabolism.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
%The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation. Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

Mathematical modeling has long contributed to our understanding of metabolism.
Decades before the genomics revolution, mechanistically structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction~\cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble and the best-fit parameter set correctly described the training data.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using unconstrained sequence-specific flux balance analysis.
The CAT yield of the model was 25\% of the maximum theoretical yield, suggesting that CAT production could be further optimized.
We also calculated CAT carbon yield for a case with transcription and translation constrained by literature; the CAT yield of the model was 38\% of this theoretical physiological case.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis itself, as well as certain reactions in amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate pathway and oxidative phosphorylation.
Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
%: mostly amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate and Entner-Doudoroff pathways and the reactions involving the energy species and cofactors.
CAT production was also robust to allosteric control, but other metabolites, specifically organic acid intermediates, were not.
Taken together, we present the first dynamic model of \textit{E. coli} cell-free protein synthesis.
We have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
We also defined the set with the lowest cost function as our best-fit set.

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements were not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).

@@

The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation; and 1 pyrophosphatase.
The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
In addition, a handful of alanine synthesis pairwise sensitivities are high.
However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO$_4$ transport.
This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
This is also likely the reason that reactions in section J rank above those in section K, despite having no effect themselves on the model.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
We first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig.~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield.

In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
In these cases the majority of the flux continued through the Entner–Doudoroff pathway, whereas in the unconstrained case the majority of flux traveled through glycolysis.
In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
In the TX/TL constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell-free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
The decrease in acetate production is promising as a mechanism to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In this study we present an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
We used the hybrid cell-free modeling approach of Wayman et al [REFHERE], which integrates traditional kinetics and a logic-based allosteric description.
We also show that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
The theoretical maximum and TX/TL-constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
The ensemble of dynamic models predicted most of the carbon flux going through glycolysis and the TCA cycle, while FBA predicted significant flux through the Entner-Doudoroff pathway.
Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolism were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
CAT production was also very sensitive to the CAT synthesis reaction, unsurprisingly.
The allosteric control component of the hybrid modeling approach was shown as important to central carbon metabolism, but not very important to CAT production.
Taken together, this is the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.

In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, this suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance. 
The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
Knockouts of \textit{gnd} and phosphate acetyltransferase show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS. 
Another limitation to be addressed in CFPS is the transcription and translation description, since the protein of interest to be expressed is ultimately bounded by these kinetic rates.
Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest that oxidative phosphorylation is important to CAT production.
Thus, this is a potential area for improvement of CFPS performance and protein yield.
